---
layout:  post
title: "Making section plots with oce and `imagep()`"
published: true
author: "Clark Richards"
date: 2016-04-25
categories: [R, oce, section, ctd]
output:
  html_document:
    mathjax:  default
    fig_caption:  true
---

```{r init, include=FALSE}
library(oce)
data(section)
```

`section` objects in the oce package are a convenient way of storing a series of CTD casts together -- indeed, the object name derives from the common name for such a series of casts collected from a ship during a single campaign. 

In it's heart, a `section` object is really just a collection of `ctd` objects, with some other metadata. The CTD stations themselves are stored as a list of `ctd` objects in the `@data` slot, like:

```{r eval=FALSE}
library(oce)
data(section)
str(section@data$station, 1)
```

    List of 124
     $ :Formal class 'ctd' [package "oce"] with 3 slots
     $ :Formal class 'ctd' [package "oce"] with 3 slots
     $ :Formal class 'ctd' [package "oce"] with 3 slots
     $ :Formal class 'ctd' [package "oce"] with 3 slots
     $ :Formal class 'ctd' [package "oce"] with 3 slots
     $ :Formal class 'ctd' [package "oce"] with 3 slots
     $ :Formal class 'ctd' [package "oce"] with 3 slots
     $ :Formal class 'ctd' [package "oce"] with 3 slots
      [list output truncated]
    
Just to prove it, we can plot make a standard `ctd` plot of one of them, by accessing them directly with the `[[` accessor syntax. Let's plot the 100th station:
```{r plotctd}
plot(section[['station']][[100]])
```

## Making nice plots of the sections themselves

The main advantage of a `section` object is to be able to quickly make plots summarizing all the data in the section. This is accomplished using the `plot` method for `section` objects, which you can read about by doing `?"plot,section-method"`. For example, to make a contour plot of the temperature:
```{r temperature, fig.height=3.5, warning=FALSE}
plot(section, which='temperature', xtype='distance')
```

Ok, cool. But what about some colors? Use the `ztype='image'` argument!
```{r temperature2, fig.height=3.5, warning=FALSE, dev.args=list(antialias='none')}
plot(section, which='temperature', xtype='distance',
     ztype='image')
```

## Finer control over the section plot

To get finer control over the section plot than is possible with the `section` `plot()` method, one trick I will sometimes do is extract the data I want from the section as a gridded matrix, and then plot the matrix directly using the `imagep()` function.

First, we "grid" the section so that all the stations comprise the same pressure levels:
```{r grid, warning=FALSE}
s <- sectionGrid(section, p='levitus')
```

Now, we can loop through the `station` fields, extracting the data as we go.
```{r extractdata}
nstation <- length(s[['station']])
p <- unique(s[['pressure']])
np <- length(p)
T <- S <- array(NA, dim=c(nstation, np))
for (i in 1:nstation) {
    T[i, ] <- s[['station']][[i]][['temperature']]
    S[i, ] <- s[['station']][[i]][['salinity']]
}
```

Basically, what we're doing here is creating an empty matrix, then filling each row with the data from the section stations. We can make a quick plot with `imagep()`:
```{r plot1}
distance <- unique(s[['distance']])
par(mfrow=c(2, 1))
imagep(distance, p, T, col=oceColorsTemperature, flipy=TRUE)
imagep(distance, p, S, col=oceColorsSalinity, flipy=TRUE)
```

Or we can do some fancier things, like use the [colormap()](http://clarkrichards.org/r/oce/plotting/colormap/2016/04/09/using-the-oce-colormap-function/) function and plot some filled contours:
```{r plot2, dev.args=list(antialias='none')}
par(mfrow=c(2, 1))
Tcm <- colormap(T, breaks=seq(4, 24, 2), col=oceColorsTemperature)
Scm <- colormap(S, breaks=seq(34, 36.8, 0.2), col=oceColorsSalinity)
imagep(distance, p, T, colormap=Tcm, flipy=TRUE,
       ylab='p [dbar]', filledContour=TRUE,
       zlab='temperature [degC]')
imagep(distance, p, S, colormap=Scm, flipy=TRUE,
       xlab='distance [km]', ylab='p [dbar]', filledContour=TRUE,
       zlab='salinity')
```
