---
layout:  post
title: "Calculating buoyancy frequency for argo/section objects using the apply() family"
published: true
author: "Clark Richards"
date: 2016-02-29
categories: [R, oce, argo, section, ctd]
output:
  html_document:
    mathjax:  default
    fig_caption:  true
---

The most recent [CRAN release of oce](https://cran.r-project.org/web/packages/oce/) includes some nice new functionality for reading and converting `argo` objects (see [http://www.argo.ucsd.edu/](http://www.argo.ucsd.edu/) for more information about the fantastic Argo float program). One question that arose out of this increased functionality was how to calculate $$ N^2 $$ (also known as the buoyancy or Brunt-Väisälä frequency) for such objects.

## Buoyancy frequency

The definition of $$ N^2 $$ is:

$$ N^2 = \frac{-g}{\rho} \frac{\partial \rho}{\partial z} $$

where $$ g $$ is the acceleration due to gravity, $$ \rho = \rho(z) $$ is the fluid density, and $$ z $$ is the vertical coordinate. Essentially $$ N^2 $$ describes the vertical variation of fluid *density* (also known as "stratification"). 

Calculating $$ N^2 $$ for regular `ctd` objects is easily accomplished with the function `oce::swN2()`. A caution: readers are encouraged to read the documentation carefully, as the details of the actual calculation can have important consequences when applied to real ocean data.

## $$ N^2 $$ for `station` objects

For the case of a `station` object (which is essentially a collection of `ctd` stations), the most straightforward way to calculate $$ N^2 $$ is to use the `lapply()` function to "apply" the `swN2()` function to each of the stations in the object. An example:

```{r station, message=FALSE}
library(oce)
data(section)
section[['station']] <- lapply(section[['station']],
                               function(x) oceSetData(x, 'N2', swN2(x, df=100)))
```
The line with the `lapply()` command takes the list of stations from the `section` object, and evaluates each of the resulting `ctd` objects using the `oceSetData()` function to add the result of `swN2()` back into the station `@data` slot.


If we wanted to make a nice plot of the result, we could do:
```{r stationplot, fig.width=7, fig.height=3.5}
col <- colorRampPalette(c('white', rev(oceColorsViridis(32))))
plot(section, which='N2', ztype='image', zbreaks=seq(0, 1e-4, 0.5e-5), zcol=col)
```
where I've defined a custom colormap just for the fun of it. 

## $$ N^2 $$ for `argo` objects

In an `argo` object, the default storage for the profiles is a matrix, rather than a list of `ctd` objects. To calculate $$ N^2 $$ and make a plot, the simplest approach would be to use `as.section()` to convert the `argo` object to a `section` class object and then do as above. However, having the field as a matrix allows for greater flexibility in plotting, e.g. using the `imagep()` function, so one might want to calculated $$ N^2 $$ in a manner consistent with the default `argo` storage format.

Let's load some example data from the `argo` dataset included in `oce`:
```{r argo}
data(argo)
argo <- argoGrid(argo)
```
Note that I've gridded the argo fields so the matrices are at consistent pressure levels. Now we create a function that can be applied to each of the matrix columns, to calculate $$ N^2 $$ from a single column of the density matrix:
```{r argo-function}
N2 <- function(x) {
    ok <- !is.na(x)
    approx(argo[['pressure']][ok,1],
           swN2(argo[['pressure']][ok,1], x[ok], df=50),
           argo[['pressure']][,1])$y
}
```
This looks a bit complicated because of some details of the object contents and of the $$ N^2 $$ calculation. First, `swN2()` can't handle missing values in the density, so we identify them with the logical `ok`. Then, because we want the final $$ N^2 $$ profile to be at the same pressure levels as the input (but without the missing levels), we use `approx()` to interpolate. Also note that we can access the vector of pressures with `argo[['pressure']][ok,1]` because in R functions can look "up" to the parent environment for unspecified variables.

Now we use the above function `N2` to calculate buoyancy frequency and add it back to the original object, like:
```{r argo-add}
argo <- oceSetData(argo, 'N2', apply(swSigmaTheta(argo), 2, N2) )
```
Note that because of the difference between the "list" and "matrix" approach, the `oceSetData()` occurs *outside* of the `apply()`. Also note the second argument in the `apply()` call, which specifies to apply the `N2()` function along the 2nd dimension of the density matrix, i.e. along columns.

Now, lets make a sweet plot of the `N2` field using `imagep()`!
```{r argoplot, fig.width=7, fig.height=3.5}
p <- argo[['pressure']][,1]
t <- argo[['time']]
imagep(t, p, t(argo[['N2']]), flipy=TRUE, col=col, breaks=seq(0, 1e-4, 1e-6),
       ylab='pressure [dbar]',
       zlab=expression(N^2~group('[', 1/s^2, ']')), zlabPosition = 'side',
       mar=c(3, 3, 2, 0.5))
```

A thing of stratified beauty, if I do say so myself.
