---
layout:  post
title: "Discrete colorbars with R and the oce package"
published: true 
author: "Clark Richards"
date: 2019-12-28
categories: [R, oce, colormap]
output:
  html_document:
    mathjax:  default
    fig_caption:  true
---

# Introduction

Making plots in oceanography (or anything, really) often requires creating some kind of "color map" -- that is, having a color represent a field in a plot that is otherwise two-dimensional. Frequently this is done when making "image"-style plots (known in MatlabTM parlance as "pcolor" or pseudocolor plots), but could also be in coloring points on a 2D scatter plot based on a third variable (e.g. a TS plot with points colored for depth).

There are a whole bunch of different ways to make colormaps in R, including various approaches that are derived from the "tidyverse" and `ggplot2` package for analyzing and plotting data. I don't really use that approach for most of my work, so won't touch on them here.

Instead, the purpose of this post (inspired by a question from a colleague who I know is a Matlab and Python user) is to show some of the ways various functions contained in the [`oce`](https://dankelley.github.io/oce) package cab be used to make colormaps and colorbars (or "palettes"). In particular, for the case where one wants a "discrete" (i.e. not continuous) colormap.

# The `imagep()` function

For making image-style plots, the function `imagep()` provided by the `oce` package is a handy function for quickly making nice-looking pseudocolor plots of matrices. The "p" in `imagep()` stands for "palette" or "pseudocolor". Mostly, it is a wrapper around the base R function `image()`, to allow for increased control of the axes, colors, and palette specification.

```{r fig.width=10, fig.height=4}
library(oce)
library(ocedata)
library(cmocean)
data(levitus)
par(mfrow=c(1, 3))
with(levitus, imagep(longitude, latitude, SST))
with(levitus, imagep(longitude, latitude, SST, col=oceColorsJet, breaks=-2:25))
with(levitus, imagep(longitude, latitude, SST, col=cmocean('thermal'), breaks=seq(-2, 30, 2)))
```

The above example shows how to use `imagep()` with 3 different colormaps (the default, the classic "jet" scheme, and the `cmocean` package) to generate an image plot with a nice palette automatically placed on the side.

# The `drawPalette()` and `colormap()` functions

Under the hood, the `imagep()` function calls another function to actually draw the palette on the side of the plot -- the `drawPalette()` function. That function can be called on it's own, enabling plot building to be much more flexible. Additionally there is the `colormap()` function, which allows for detailed specification of the colormap properties to use, which can then be passed as an object to `drawPalette()`.

For example, say we wanted to make a TS (temperature-salinity) plot of the Levitus surface data, but with each point colored by latitude in 10 degree increments. We do that first by making a colormap object:

```{r}
cm <- colormap(expand.grid(levitus$latitude, levitus$longitude)[,1], breaks=seq(-90, 90, 10), col=oceColorsViridis)
```
(note that we use `expand.grid()` to make the number of lon/lat points match the matrices).

Looking inside the object, we can see some of the details that further plotting/palette functions can make use of:
```{r}
str(cm)
```
Some of those fields are obvious (some probably aren't to inexperienced users) but one field that is handy to know about is the `zcol` field. This encodes a color for every value in the original object based on the colormap specification. So, we can make a plot with the points colored based on the colormap using the argument `col=cm$zcol`. We can also add the palette to the plot using the `drawPalette()` function, which has to be called *before* the main plot:
```{r}
drawPalette(colormap=cm)
plotTS(with(levitus, as.ctd(SSS, SST, 0)), pch=19, col=cm$zcol, mar=par('mar'))
```

One problem that can happen when there are a lot of points is that overplotting obscures patterns in the colors. An easy way to fix this is to randomize the order of the plotted points with the `sample()` function:
```{r}
drawPalette(colormap=cm)
sI <- sample(seq_along(cm$zcol))
plotTS(with(levitus, as.ctd(SSS[sI], SST[sI], 0)), pch=19, col=cm$zcol[sI], mar=par('mar'))
```

## Custom color palettes with `colormap()`

In addition to the "known" color palettes that are included in R and `oce` (see also the `cmocean` package below, as well as `RcolorBrewer`), the `colormap()` function has arguments that allow for custom-built palettes. Specifically the `x0`, `x1`, `col0` and `col1` arguments, which are detailed in the help file as:

```
x0, x1, col0, col1: Vectors that specify a color map.  They must all be
          the same length, with ‘x0’ and ‘x1’ being numerical values,
          and ‘col0’ and ‘col1’ being colors.  The colors may be
          strings (e.g. ‘"red"’) or colors as defined by rgb or
          hsv.
```

The idea is that the `x0` values define the numeric level of the *bottom* of the color ranges, the `x1` values define the *top* of the color ranges, and the `col0` and `col1` the colors associated with the levels. An example:

```{r}
cm <- colormap(expand.grid(levitus$latitude, levitus$longitude)[,1],
               x0=c(-90, -45, 0, 45), x1=c(-45, 0, 45, 90),
               col0=c(1, 2, 3, 4), col1=c(2, 3, 4, 5))
drawPalette(colormap=cm)
sI <- sample(seq_along(cm$zcol))
plotTS(with(levitus, as.ctd(SSS[sI], SST[sI], 0)), pch=19, col=cm$zcol[sI], mar=par('mar'))
```

(Note that to make the above plot, I had to fix a [bug](https://github.com/dankelley/oce/pull/1644) in `oce` that was making the `zcol` come out as "black" for all cases. Either build oce from source, or wait for the update to get pushed to CRAN in a month or two).

