---
layout:  post
title: "Negating functions and function definitions: an 'opposite' function to the wonderful `%in%` operator"
published: true
author: "Clark Richards"
date: 2019-11-13
output:
  html_document:
    mathjax:  default
    fig_caption:  true
---

## Introduction

R has some neat functions, and even some weird quirks, that you aren't likely to discover on your own but can either be immensely helpful or horribly confounding. 

For example, the "+" operator (i.e. addition) is actually a *function*, and can even be called using the typical "bracket" notation:
```{r}
1 + 2
```
We can use backticks to evaluate the function as a "regular" function:
```{r}
`+`
```
And can therefore call it as a "regular" function, using brackets to pass the arguments:
```{r}
`+`(1, 2)
```

One consequence of this is that it is possible to redefine how "+" works:
```{r}
`+` <- function(x1, x2) x1 - x2
1 + 2
```

Ok ... admittedly that's confusing. Why would you want to redefine "+"? Well, one example is given by the syntax of the [`ggplot2`](https://ggplot2.tidyverse.org/) package, which defines it's own version of "+" that lets you string plotting functions together to build up a plot (e.g. see my post about plotting [here]({% post_url 2017-12-11-recording-plots %})).

## The `%in%` function

The `%in%` function is one of those functions that just *clicked* when I started using R. It's an elegant way to write conditional statements -- it checks whether the object to the left of the operator occurs anywhere in the object on the right of the operator. An example:

Does the number 5 occur in the vector of 1 through 10?
```{r}
5 %in% 1:10
```
Yes, it does (obviously).

One thing that I often find myself doing however, is wanting to know if something *doesn't* occur in another object. To make that work, I usually wrap the whole statement in brackets and then precede with an `!` operator (logical negation). Like this:
```{r}
!(11 %in% 1:10)
```
This evaluates to `TRUE` because 11 is **not** in the vector 1:10. While this works, it's always bugged me because it just looks inelegant.

Well, while browsing Twitter recently, I came across this post from [@groundwalkergmb](https://twitter.com/groundwalkergmb):

<blockquote class="twitter-tweet"><p lang="en" dir="ltr">Am I the only one that always does <br><br>if(!(blah %in% woo)) <br><br>instead of <br><br>if(!blah %in% woo) <br><br>because the latter makes me really uncomfortable even though it works fine (and equivalently to the former)? <a href="https://twitter.com/hashtag/rstats?src=hash&amp;ref_src=twsrc%5Etfw">#rstats</a></p>&mdash; Gabe Becker (@groundwalkergmb) <a href="https://twitter.com/groundwalkergmb/status/1194698809760108544?ref_src=twsrc%5Etfw">November 13, 2019</a></blockquote> <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>

First, note the (confusing) syntax that you don't actually have to negate the entire expression. It is equivalent to write:
```{r}
!11 %in% 1:10
```
Ew. Gross! (@groundwalkergmb agrees with me about this)

But reading through the comments led me to this one:

<blockquote class="twitter-tweet"><p lang="en" dir="ltr">`%nin% &lt;- Negate(`%in%`)<br><br>if(blah %nin% woo) {}</p>&mdash; Brandon Hurr (@bhive01) <a href="https://twitter.com/bhive01/status/1194703071193878528?ref_src=twsrc%5Etfw">November 13, 2019</a></blockquote> <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>

Basically, you can use the `Negate()` function (never knew about this before) to create a *new* function which returns the logical negation of the output of the original.
```{r}
`%nin%`<- Negate(`%in%`)
11 %nin% 1:10
```

<blockquote class="twitter-tweet" data-conversation="none"><p lang="en" dir="ltr">What?!? <a href="https://t.co/24udunAkgV">pic.twitter.com/24udunAkgV</a></p>&mdash; Clark Richards, PhD (@ClarkGRichards) <a href="https://twitter.com/ClarkGRichards/status/1194756668774658048?ref_src=twsrc%5Etfw">November 13, 2019</a></blockquote> <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>
