---
layout:  post
title: "A perfect CTD profile"
published: true
author: "Clark Richards"
date: 2019-11-14
output:
  html_document:
    mathjax:  default
    fig_caption:  true
---

I love the $$\tanh$$ function. A lot. It's such a perfect model for a
density interface in the ocean, that it is commonly used in
theoretical and numerical models and I regularly used it for both
research and demonstration/example purposes. Behold, a $$\tanh$$ interface:

$$ T(z) = T_0 + \delta T \tanh \left( \frac{z-z_0}{dz} \right) $$

```{r}
T <- function(z, T0=10, dT=5, z0=-25, dz=5) T0 + dT*tanh((z - z0)/dz)
z <- seq(0, -60)
plot(T(z), z)
```

But whenever I use it, especially for teaching, I'm always saying how it's idealized and really doesn't represent what an ocean interface actually looks like. UNTIL NOW.

```{r message=FALSE, warnings=FALSE}
library(oce)
ctd <- read.oce('../assets/D19002034.ODF')
par(mfrow=c(1, 2))
plot(ctd, which=1, type='l')
plot(ctd, which=5)
```

Yes, this is real data.

Just how close to a $$\tanh$$ is it?[^dan]
```{r}
z <- -ctd[["depth"]]
T <- ctd[["temperature"]]
m <- nls(T~a+b*tanh((z-z0)/dz), start=list(a=3, b=1, z0=-10, dz=5))
m
plot((T-predict(m))/T * 100, z, type="o", xlab='Percent error')
```

[^dan]: My PhD advisor, who also taught me introductory physical oceanography, once said to a class of students while using tanh to describe an idealized interface: "tanh -- it's like 'lunch', only better!"
