---
layout:  post
title: "Using the oce package to make nice maps"
published: true
author: "Clark Richards"
date: 2019-07-12
categories: [oce]
output:
  html_document:
    mathjax:  default
    fig_caption:  true
---

## Introduction

Making maps is a pretty important part of doing, and presenting, ocean data analyses. Except for very small domains, using map projections is crucial to ensure that the map is not distorted. This is particularly true for polar and high latitude regions, such as the Arctic (where I do much of my work).

In this post I will give a brief introduction to making projected maps with the `oce` package, including not just the land/coastline but also various ways of plotting the bathymetry. For the latter I will discuss the handy [`marmap` package](https://cran.rstudio.com/web/packages/marmap/index.html).

## Making a polar stereographic map of the Canadian Arctic

To make "projected" maps with `oce`, you can use the `mapPlot()` function. The easiest way is to use `mapPlot()` with a `coastline` object, to set the projection you want, and then add other elements (such as bathymetry, etc) using functions like `mapLines()`, `mapPoints()`, `mapImage()`, etc. You may need to redo the coastline at the end to clean up anything that might have plotted over land that you don't want.

The `coastlineWorldFine` dataset from the `ocedata` package is pretty good for sub-regions such as the Canadian Arctic.

```{r}
library(oce)
library(ocedata) #for the coastlineWorldFine data
data(coastlineWorldFine)
```

To get the projection you want, it must be passed in the `projection=` argument using "proj4" syntax as a character string. You can read up on the syntax and available projection in the help -- i.e. `?mapPlot`, or have a look at:

[PROJ4]https://proj.org/operations/projections/index.html)

For polar maps, the most commonly used is the stereoraphic projection:

```{r}
## Save it to a function to make it easy to re-run
mp <- function() {
    mapPlot(coastlineWorldFine, projection="+proj=stere +lon_0=-90 +lat_0=90",
            longitudelim = c(-120, -60),
            latitudelim = c(60, 85), col='grey')
}
mp()
```

In this example, the `+lon_0` parameter defines the longitude at the center of the projection, and the `+lat_0` is the latitude at the center of the projection. The `mapPlot()` arguments `longitudelim` and `latitudelim` are used to control the extent of the map.

As for adding bathymetry, there are a few options:

1. `oce` includes a low-res version of the etopo dataset, called `topoWorld` that can be used, either as a contour plot or as an image plot.

2. The `marmap` package allows easy downloading of subsets of the full resolution etopo data, which can be used to make nicer plots. One downside is that because of the "triangle" nature of a polar stereoraphic projection you have to download a lot more data than you really need.

### Using `topoWorld`

A quick way of adding bathymetry is to simply use the `mapContour()` function (analogous to the base-R `contour()`).

```{r}
data(topoWorld)
mp()
mapContour(topoWorld, levels=-c(500, 1000, 2000, 4000),
           drawlabels=FALSE)
```

Contour plots can be tricky, because of the labeling etc, and choosing which contours to plot. Another option is to use the `mapImage()` function:

```{r}
mp()
mapImage(topoWorld, col=oceColorsGebco, breaks=seq(-4000, 0, 500))
mapPolygon(coastlineWorldFine, col='grey')
mapGrid()
```

But sometimes using `mapImage()` shows off the "blockiness" of the low-res `topoWorld` dataset -- especially near the poles where the shape of the cells (evenly divided in lon/lat space) gets really skinny.

One fix is to use a higher res data set (so the "boxes" are smaller -- see next section). Another is to use the `fillContour` argument in `mapImage()` to plot filled contours rather than the individual grid cells.

```{r}
mp()
mapImage(topoWorld, col=oceColorsGebco, breaks=seq(-4000, 0, 500), filledContour = TRUE)
mapPolygon(coastlineWorldFine, col='grey')
mapGrid()
```

### Using marmap


Here we load the `marmap` package and download the bathymetry. Note that because we have the North Pole in view, we basically have to download an entire half hemisphere. It takes a little while to download, but if you use the `keep=TRUE` argument then it will just reload it from the file each time.

However, because there is so much more data, the plotting can be quite a bit slower.

```{r}
library(marmap)
b <- as.topo(getNOAA.bathy(-180, 0, 55, 90, keep=TRUE))

mp()
mapImage(b, col=oceColorsGebco, breaks=seq(-4000, 0, 500))
mapPolygon(coastlineWorldFine, col='grey')
mapGrid()
```

Looks pretty good, though!
