---
layout:  post
title: "A new default colormap for the `oce` package"
published: true 
author: "Clark Richards"
date: 2020-08-17
categories: [R, oce, colormap]
output:
  html_document:
    mathjax:  default
    fig_caption:  true
---

# Introduction

A lot has happened since my [last post]({% post_url 2019-12-28-discrete-colorbars-with-r-and-oce %}) (and really, I mean A LOT). However, in the spirit of that last post, one good thing that has happened is that in the upcoming 1.3-0 release of the [`oce` package](https://CRAN.R-project.org/package=oce) on CRAN we have changed the default colormap (or "palette" as it's often referred in the R world) to something not nearly sucky as the classic ["jet" colormap](https://jakevdp.github.io/blog/2014/10/16/how-bad-is-your-colormap/) (originally made popular by Matlab). 

The purpose of this post is just to highlight the difference between the old and new defaults, and also to show many other great colormaps for plotting oceanographic data.

# Jet vs viridis

When the `imagep()` function was first added to `oce`, there was a limited selection of palettes that could be used. For a long time, the default of the `imagep()` function itself has been a palette called `oceColorsPalette()` (which is a blue-through-white-to-red palette most useful for diverging colormaps), however many of the internal functions that use colors defaulted to the `oceColorsJet()` palette, including: the `plot,section-method`, and more importantly, the `colormap()` function itself (which I wrote about [here]({% post_url 2016-04-09-using-the-oce-colormap-function %})).

```{r fig.height=3, fig.width=10}
library(oce)
data(volcano)
par(mfrow=c(1, 3))
imagep(volcano, col=oceColorsPalette, main='oceColorsPalette()')
imagep(volcano, col=oceColorsJet, main='oceColorsJet')
imagep(volcano, col=oceColorsViridis, main='oceColorsViridis()')
```

The palette on the left, `oceColorsPalette()` has been retained as the default for the `plot,adp-method` (is velocities are typically diverging). The palette in the middle, Jet, is still available but is no longer the default for any plots. The `viridis` palette on the right is now the default for all palette-related functions.

I won't go into the details of why `viridis` is so much better -- a quick google will turn up lots of articles (though this [new paper](https://ieeexplore.ieee.org/document/9167329) from [Kristen Thyng](http://kristenthyng.com/) has lots of great info.

# Other, even better palettes

Kristen Thyng is author the author of the `cmocean` series of oceanographic colormaps, originally created for python/matplotlib. There is now a [`cmocean`](https://CRAN.R-project.org/package=cmocean) package on CRAN, which makes all the python colormaps available in R.

```{r}
library(ocedata)
library(cmocean)
data(levitus)
par(mfrow=c(1, 2))
with(levitus, imagep(longitude, latitude, SST, col=cmocean('thermal'), main="sea surface temperature"))
with(levitus, imagep(longitude, latitude, SST, col=cmocean('haline'), main="sea surface salinity"))
```

There is also the [`viridis` package](https://cran.rstudio.com/web/packages/viridis/index.html) itself, which makes the series of palettes developed for matplotlib available in R.

```{r}
library(viridis)
imagep(volcano, col=inferno)
```
